---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
# Load libraries
library(ggplot2)
library(dplyr)
library(tidyr)

```

## Data selection from SEER



Dataset was saved as "SEER_dataset.csv"

## Download CSV file
> setwd("C:/Users/piman/OneDrive - Khon Kaen University/Desktop/SEER_EC")

```{r}
# Read the CSV file
# Interactively select the file
file_path <- file.choose()
SEER_EC <- read.csv(file_path, stringsAsFactors = FALSE)

# 43764 patients
```


## Data exploratory
```{r}
str(SEER_EC)
```


```{r}
# Function to calculate distinct values and their percentages for a column
calculate_percentage <- function(column) {
  total_count <- nrow(SEER_EC)
  column_summary <- SEER_EC %>%
    group_by(across(all_of(column))) %>%
    summarise(count = n()) %>%
    mutate(percentage = (count / total_count) * 100)
  return(column_summary)
}

# Apply the function to each column and store the results in a list
column_percentages <- lapply(names(SEER_EC), calculate_percentage)

# Display the results for each column
for (i in 1:length(column_percentages)) {
  cat("Column:", names(SEER_EC)[i], "\n")
  print(column_percentages[[i]])
  cat("\n")
}
```

```{r}
names(SEER_EC) <- c(
  "ID",                                        # Patient.ID
  "Age",                                       # Age.recode.with.single.ages.and.90.
  "Diag_Year",                                 # Year.of.diagnosis
  "Race",                                      # Race.recode..W..B..AI..API.
  "Sex",                                       # Sex
  "Hispanic",                                  # Origin.recode.NHIA..Hispanic..Non.Hisp.
  "MaritalStatus",                             # Marital.status.at.diagnosis
  "Income",                                    # Median.household.income.inflation.adj.to.2022
  "RuralUrban",                                # Rural.Urban.Continuum.Code
  "SurvivalMonths",                            # Survival.months
  "CauseSpecificDeath",                        # SEER.cause.specific.death.classification
  "OtherCauseDeath",                           # SEER.other.cause.of.death.classification
  "Vital",                                     # Vital.status.recode..study.cutoff.used.
  "SequenceNumber",                            # Sequence.number
  "PrimarySite_code",                          # Primary.Site...labeled
  "Grade",                                     # Grade.Recode..thru.2017.
  "SummaryStage",                              # SEER.Combined.Summary.Stage.2000..2004.2017.
  "HistologicBehavior",                        # ICD.O.3.Hist.behav..malignant
  "AJCC_T",                                    # Derived.AJCC.T..6th.ed..2004.2015.
  "AJCC_N",                                    # Derived.AJCC.N..6th.ed..2004.2015.
  "AJCC_M",                                    # Derived.AJCC.M..6th.ed..2004.2015.
  "TumorSize",                                 # Tumor.Size.Over.Time.Recode..1988..
  "NodesExamined",                             # Regional.nodes.examined..1988..
  "NodesPositive",                             # Regional.nodes.positive..1988..
  "TumorSizeCS",                               # CS.tumor.size..2004.2015.
  "ExtensionCS",                               # CS.extension..2004.2015.
  "LymphNodesCS",                              # CS.lymph.nodes..2004.2015.
  "MetsAtDXCS",                                # CS.mets.at.dx..2004.2015.
  "InSituMalignantTumorsTotal",                # Total.number.of.in.situ.malignant.tumors.for.patient
  "BenignTumorsTotal",                         # Total.number.of.benign.borderline.tumors.for.patient
  "TimeToTreatment",                           # Time.from.diagnosis.to.treatment.in.days.recode
  "SurgicalTreatment",                         # RX.Summ..Surg.Prim.Site..1998..
  "ScopeRegionalLNSurgery",                    # RX.Summ..Scope.Reg.LN.Sur..2003..
  "SurgicalRadiationSequence",                 # RX.Summ..Surg.Rad.Seq
  "Radiation",                                 # Radiation.recode
  "Chemotherapy",                              # Chemotherapy.recode..yes..no.unk.
  "SystemicSurgerySequence",                   # RX.Summ..Systemic.Sur.Seq..2007..
  "PrimarySite"                                # Primary.Site
)

```

## Recode data

```{r}
# Replace 'years' and any non-numeric characters, then convert to integer
SEER_EC$Age <- as.integer(gsub("[^0-9]", "", SEER_EC$Age))

# For '90+ years', you might want to decide on a specific age to assign
# For example, if you choose to assign 90 to '90+ years':
SEER_EC$Age[SEER_EC$Age == 90] <- 90
```


```{r}
# Recode 'Race' variable
SEER_EC$Race <- ifelse(SEER_EC$Race == "White", "W",
                       ifelse(SEER_EC$Race == "Black", "B", "Other"))
```


```{r}
# Recode 'Sex' variable
SEER_EC$Sex <- ifelse(SEER_EC$Sex == "Male", 1, 0)
```


```{r}
# Recode 'Hispanic' variable for Hispanic identification
SEER_EC$Hispanic <- ifelse(SEER_EC$Hispanic == "Spanish-Hispanic-Latino", 1, 0)
```


```{r}
# Recode 'MaritalStatus' variable
SEER_EC$MaritalStatus <- ifelse(SEER_EC$MaritalStatus == "Married (including common law)", 1, 0)
```


```{r}
# Recode 'Income' variable; Low: Less than $60,000, Medium: $60,000 to $99,999, High: $100,000 and above
SEER_EC <- SEER_EC %>%
  mutate(Income = case_when(
    grepl("< $40,000", Income) | grepl("^\\$40,000 - \\$44,999", Income) | grepl("^\\$45,000 - \\$49,999", Income) |
    grepl("^\\$50,000 - \\$54,999", Income) | grepl("^\\$55,000 - \\$59,999", Income) ~ "Low",
    grepl("^\\$60,000 - \\$64,999", Income) | grepl("^\\$65,000 - \\$69,999", Income) | 
    grepl("^\\$70,000 - \\$74,999", Income) | grepl("^\\$75,000 - \\$79,999", Income) |
    grepl("^\\$80,000 - \\$84,999", Income) | grepl("^\\$85,000 - \\$89,999", Income) |
    grepl("^\\$90,000 - \\$94,999", Income) | grepl("^\\$95,000 - \\$99,999", Income) ~ "Medium",
    grepl("^\\$100,000 - \\$109,999", Income) | grepl("^\\$110,000 - \\$119,999", Income) | 
    grepl("^\\$120,000\\+", Income) ~ "High",
    TRUE ~ "Medium"  # Default to Medium if none of the above match
  ))
```


```{r}
# Recode 'RuralUrban' variable
SEER_EC <- SEER_EC %>%
  mutate(RuralUrban = case_when(
    grepl("metropolitan", RuralUrban, ignore.case = TRUE) ~ 1,
    grepl("Nonmetropolitan", RuralUrban, ignore.case = TRUE) ~ 0,
    TRUE ~ 0  # Default to 0 if none of the above match
  ))
```


```{r}
# Convert 'SurvivalMonths' from character to integer
SEER_EC$SurvivalMonths <- as.integer(SEER_EC$SurvivalMonths)
```


```{r}
# Recode 'SequenceNumber' variable
SEER_EC <- SEER_EC %>%
  mutate(SequenceNumber = case_when(
    SequenceNumber %in% c("One primary only", "1st of 2 or more primaries") ~ 1,
    TRUE ~ 0  # Keeps all other values as 0
  ))
```


```{r}
# Recode 'PrimarySite_code' variable

# Only remove text after a dash, if it exists, while keeping the code before the dash intact
SEER_EC$PrimarySite_code <- gsub("-.*", "", SEER_EC$PrimarySite_code)

# Replace '.' with '_' to meet the format requirement 
SEER_EC$PrimarySite_code <- gsub("\\.", "_", SEER_EC$PrimarySite_code)

```


```{r}
# Recode 'Grade' variable
SEER_EC <- SEER_EC %>%
  mutate(Grade = case_when(
    grepl("Grade I$", Grade) ~ "Grade I",                       # Ensures it ends with "Grade I"
    grepl("Grade II$", Grade) ~ "Grade II",                     # Ensures it ends with "Grade II"
    grepl("Grade III$", Grade) ~ "Grade III",                   # Ensures it ends with "Grade III"
    grepl("Grade IV$", Grade) ~ "Grade IV",                     # Ensures it ends with "Grade IV"
    TRUE ~ "Unknown"                                           # Handles all other cases
  ))
```


```{r}
# Recode 'SummaryStage' variable
SEER_EC <- SEER_EC %>%
  mutate(SummaryStage = case_when(
    SummaryStage == "Distant site(s)/node(s) involved" ~ "Dist+LN",
    SummaryStage == "Unknown/unstaged/unspecified/DCO" ~ "unknown",
    SummaryStage == "Localized only" ~ "Local",
    SummaryStage == "Regional by direct extension only" ~ "Regi+Ext",
    SummaryStage == "Regional lymph nodes involved only" ~ "Reg+LN",
    SummaryStage == "Regional by both direct extension and lymph node involvement" ~ "Reg+Ext/LN",
    TRUE ~ SummaryStage  # Preserves the original value in case of non-matching cases
  ))
```


```{r}
# Recode 'HistogoicBehavior' variable
SEER_EC <- SEER_EC %>%
  mutate(HistologicBehavior = case_when(
    grepl("^8000/3:", HistologicBehavior) ~ "Neoplasm",
    grepl("^8010/3:", HistologicBehavior) ~ "Carcinoma",
    grepl("^8070/3:", HistologicBehavior) ~ "SCC",
    grepl("^8071/3:", HistologicBehavior) ~ "SCC_keratinizing",
    grepl("^8140/3:", HistologicBehavior) ~ "Adenocarcinoma",
    grepl("^8490/3:", HistologicBehavior) ~ "SRC_carcinoma",
    TRUE ~ "Other"  # Handles all other cases
  ))
```


```{r}
# Recode 'TumorSizeCS' variable
SEER_EC <- SEER_EC %>%
  mutate(
    TumorSizeCS_polyposis = ifelse(TumorSizeCS == 998, 1, 0),  # Create new column based on condition
    TumorSizeCS = case_when(
      TumorSizeCS %in% c(999, 888) ~ NA_real_,  # Set 999, and 888, as NA (missing data)
      TumorSizeCS == 998 ~ 0,
      TumorSizeCS == 990 ~ 10,
      TumorSizeCS == 991 ~ 90,
      TumorSizeCS == 992 ~ 190,
      TumorSizeCS == 993 ~ 290,
      TumorSizeCS == 994 ~ 390,
      TumorSizeCS == 995 ~ 490,
      between(TumorSizeCS, 1, 10) ~ 10,
      between(TumorSizeCS, 11, 20) ~ 20,
      between(TumorSizeCS, 21, 30) ~ 30,
      between(TumorSizeCS, 31, 40) ~ 40,
      between(TumorSizeCS, 41, 50) ~ 50,
      between(TumorSizeCS, 51, 60) ~ 60,
      between(TumorSizeCS, 61, 70) ~ 70,
      between(TumorSizeCS, 71, 80) ~ 80,
      between(TumorSizeCS, 81, 90) ~ 90,
      between(TumorSizeCS, 91, 100) ~ 100,
      TumorSizeCS >= 101 & TumorSizeCS <= 989 ~ 10 * ceiling(TumorSizeCS / 10),  # General case for values 101 to 988
      TRUE ~ TumorSizeCS  # Catch-all in case something is outside expected range
    )
  )
```


```{r}
# Recode 'ExtensionCS' variable
SEER_EC <- SEER_EC %>%
  mutate(ExtensionCS = ifelse(ExtensionCS == 999, NA_real_, ExtensionCS))
```


```{r}
# Recode 'LymphNodesCS' variable
SEER_EC <- SEER_EC %>%
  mutate(LymphNodesCS = ifelse(LymphNodesCS == 999, NA_real_, LymphNodesCS))
```


```{r}
# Recode 'MetsAtDXCS' variable
SEER_EC <- SEER_EC %>%
  mutate(MetsAtDXCS = ifelse(MetsAtDXCS == 99, NA_real_, MetsAtDXCS))
```

ExtensionCS/LymphnodesCS/MetsAtDXCS values refer to specific codes which nor represent "continuous variable".
Consider to encode the variables before running ML

```{r}
# Recode 'SurgicalTreatment' variable
SEER_EC <- SEER_EC %>%
  mutate(SurgicalTreatment = case_when(
    SurgicalTreatment == 99 ~ NA_character_,   # Set 99 as NA (missing data)
    SurgicalTreatment == 0 ~ "No_surgery",
    SurgicalTreatment == 30 ~ "Partial_esophagectomy",
    SurgicalTreatment == 40 ~ "Total_esophagectomy",
    SurgicalTreatment == 80 ~ "Esophagectomy",
    SurgicalTreatment == 90 ~ "Surgery",
    between(SurgicalTreatment, 10, 19) ~ "Local_destruction",
    between(SurgicalTreatment, 20, 29) ~ "Local_excision",
    between(SurgicalTreatment, 50, 60) ~ "Esophagectomy+",
    TRUE ~ as.character(SurgicalTreatment)  # Keeps other values as they are, converted to character for consistency
  ))
```


```{r}
# Recode 'ScopeRegionalLNSurgery' variable
SEER_EC <- SEER_EC %>%
  mutate(ScopeRegionalLNSurgery = case_when(
    ScopeRegionalLNSurgery == "None" ~ "0",  # Change "None" to 0
    ScopeRegionalLNSurgery == "Unknown or not applicable" ~ NA_character_,  # Set "Unknown or not applicable" as NA
    TRUE ~ "1"  # Change all other values to 1
  ))
```


```{r}
# Recode 'SurgicalRadiationSequence' variable
SEER_EC <- SEER_EC %>%
  mutate(SurgicalRadiationSequence = case_when(
    SurgicalRadiationSequence %in% c("Radiation prior to surgery", "Radiation before and after surgery") ~ "Rad_before",
    SurgicalRadiationSequence %in% c("Radiation after surgery", "Surgery both before and after radiation") ~ "Rad_after",
    TRUE ~ "Other"  # Sets all other values to "Other"
  ))
```


```{r}
# Recode 'Radiation' variable
SEER_EC <- SEER_EC %>%
  mutate(Radiation = case_when(
    Radiation %in% c("None/Unknown", "Refused (1988+)") ~ "0",
    TRUE ~ "1"
  ))
```


```{r}
# Recode 'Chemotherapy' variable
SEER_EC <- SEER_EC %>%
  mutate(Chemotherapy = ifelse(Chemotherapy == "Yes", "1", "0"))

# 43764 patients
```

## Exclude data

```{r}
# Filter "OtherCauseDeath"
SEER_EC <- SEER_EC %>%
  filter(!OtherCauseDeath %in% c("Dead (attributable to causes other than this cancer dx)", 
                                 "Dead (missing/unknown COD)"))

# 36790 patients
```


```{r}
# Remove rows where SurvivalMonths < 60 and Vital is "Alive"
SEER_EC <- SEER_EC %>%
  filter(!(SurvivalMonths < 60 & Vital == "Alive"))

# 36494 patients
```



```{r}
# Filtering rows where SurvivalMonths is less than 1 month
SEER_EC <- SEER_EC %>%
  filter(SurvivalMonths >= 1)

# 33128 patients
```


```{r}
# Check for duplicates in the 'ID' column
duplicates_in_ID <- SEER_EC[duplicated(SEER_EC$ID) | duplicated(SEER_EC$ID, fromLast = TRUE), ]

# Print the duplicates
if (nrow(duplicates_in_ID) > 0) {
    print("Duplicate IDs found:")
    print(duplicates_in_ID)
} else {
    print("No duplicate IDs found.")
}
```

```{r}
# Remove duplicate IDs, keeping the first occurrence of each ID
SEER_EC <- SEER_EC %>%
  group_by(ID) %>%
  filter(row_number() == 1) %>%
  ungroup()

# 33105 patients
```


```{r}
# Filter the dataframe for ages between 20 and 80
SEER_EC <- SEER_EC %>%
  filter(Age >= 20 & Age <= 80)

# 28478 patients
```


```{r}
# Remove rows with missing data from SEER_EC
SEER_EC <- na.omit(SEER_EC)

# 13896 patients
```


## Create dataframe

```{r}
# Create EC_df by selecting the columns from SEER_EC
EC_df <- SEER_EC %>%
  select(ID, 
         SurvivalMonths,
         Diag_Year,
         Age, 
         Race, 
         Sex, 
         Hispanic, 
         MaritalStatus, 
         Income, 
         SequenceNumber, 
         PrimarySite_code, 
         Grade, 
         SummaryStage, 
         HistologicBehavior, 
         AJCC_T, 
         AJCC_N, 
         AJCC_M, 
         TumorSizeCS, 
         TumorSizeCS_polyposis,
         ExtensionCS, 
         LymphNodesCS, 
         MetsAtDXCS, 
         InSituMalignantTumorsTotal, 
         BenignTumorsTotal, 
         SurgicalTreatment, 
         ScopeRegionalLNSurgery, 
         SurgicalRadiationSequence, 
         Radiation, 
         Chemotherapy, 
         )
```


## Save dataframe

```{r}
# Save the cleaned EC_df dataframe to a CSV file
write.csv(EC_df, "EC_df.csv")
```


```{r}
# Function to calculate distinct values and their percentages for a column
calculate_percentage <- function(column) {
  total_count <- nrow(EC_df)
  column_summary <- EC_df %>%
    group_by(across(all_of(column))) %>%
    summarise(count = n()) %>%
    mutate(percentage = (count / total_count) * 100)
  return(column_summary)
}

# Apply the function to each column and store the results in a list
column_percentages <- lapply(names(EC_df), calculate_percentage)

# Display the results for each column
for (i in 1:length(column_percentages)) {
  cat("Column:", names(EC_df)[i], "\n")
  print(column_percentages[[i]])
  cat("\n")
}

```


```{r}
# Plotting the histogram of SurvivalMonths
ggplot(EC_df, aes(x = SurvivalMonths)) +
  geom_histogram(aes(y = ..count..), binwidth = 5, color = "black", fill = "blue", alpha = 0.6) +
  labs(title = "Distribution of Survival Months",
       x = "Survival Months",
       y = "Frequency") +
  theme_minimal()
```

```{r}
# Calculate density data
density_data <- density(EC_df$SurvivalMonths)

# Convert density data to a data frame
density_df <- data.frame(SurvivalMonths = density_data$x, Density = density_data$y)

# Plot with histogram, density, and colored areas based on the cutoff
ggplot(EC_df, aes(x = SurvivalMonths)) +
  geom_histogram(aes(y = ..count..), binwidth = 4, color = "grey3", fill = "grey", alpha = 0.5) +
  geom_area(data = subset(density_df, SurvivalMonths < 60),
            aes(x = SurvivalMonths, y = Density * length(EC_df$SurvivalMonths) * 5), 
            fill = "#FA8072", alpha = 0.5) +
  geom_area(data = subset(density_df, SurvivalMonths >= 60),
            aes(x = SurvivalMonths, y = Density * length(EC_df$SurvivalMonths) * 5), 
            fill = "#008080", alpha = 0.5) +
  geom_vline(xintercept = 60, color = "grey9", linetype = "dashed", size = 0.8) +
  annotate("text", x = 60*1.2, y = max(density_df$Density * length(EC_df$SurvivalMonths) * 5), 
           label = "60 months", vjust = -0.5, color = "black") +
  labs(x = "Survival months",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text = element_text(size = 12), # Increase tick label font size
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  xlim(-10, 150)
```


```{r}
# Ensure Diag_Year is a factor for better categorical plotting
EC_df$Diag_Year <- as.factor(EC_df$Diag_Year)

# Plot the data using a boxplot without displaying outliers
ggplot(EC_df, aes(x = Diag_Year, y = SurvivalMonths, fill = Diag_Year)) +
  geom_boxplot(outlier.color = NA) +  # Exclude outliers by setting outlier color to NA
  scale_fill_viridis_d(guide = FALSE) +  # Use a discrete color scale and disable the legend
  labs(title = "Boxplot of Survival Months by Diagnosis Year",
       x = "Diagnosis Year",
       y = "Survival Months") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```


```{r}
# Calculate the median and standard error of the median for each diagnosis year
median_survival_by_year <- SEER_EC %>%
  group_by(Diag_Year) %>%
  summarise(
    MedianSurvival = median(SurvivalMonths, na.rm = TRUE),
    SEM = sd(SurvivalMonths, na.rm = TRUE) / sqrt(n())  # Calculating SEM
  ) %>%
  mutate(
    LowerBound = MedianSurvival - SEM,
    UpperBound = MedianSurvival + SEM
  )

# Print the data to check SEM and bounds
print(median_survival_by_year)

# Plot the data with a ribbon for error interval
ggplot(median_survival_by_year, aes(x = as.numeric(Diag_Year), y = MedianSurvival)) +
  geom_line(color = "blue") +  # Line plot
  geom_point(color = "red") +  # Points
  geom_ribbon(aes(ymin = LowerBound, ymax = UpperBound), fill = "blue", alpha = 0.4) +  # Ribbon for error interval
  labs(title = "Median Survival Months by Diagnosis Year with Confidence Band",
       x = "Diagnosis Year",
       y = "Median Survival Months") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```


```{r}
# Calculate the median SurvivalMonths for each age
age_survival_stats <- EC_df %>%
  group_by(Age) %>%
  summarise(MedianSurvival = median(SurvivalMonths, na.rm = TRUE))

# View the resulting data to ensure correctness
print(age_survival_stats)

# Plotting the median SurvivalMonths by age
ggplot(age_survival_stats, aes(x = Age, y = MedianSurvival)) +
  geom_point(color = "blue") +  # Scatter plot points
  geom_line(group = 1, color = "red") +  # Connect points with a line
  labs(title = "Median Survival Months by Age",
       x = "Age",
       y = "Median Survival Months") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```


```{r}
```

